---
author: "Evan Krook"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, fig.height = 4)
library(tidyverse)
library(lubridate)
library(scales)
library(modelr)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\renewcommand{\prob}{\mathsf{P}}
\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\SD}{\mathsf{SD}}
\newcommand{\SE}{\mathsf{SE}}

## Homework Assignment 10

#### Due Friday, December 2, 2022, at 11:59 PM

## Problems

The *dugong.csv* data set contains data on 27 dugongs, which are marine mammals.  Since we cannot ask a dugong how old it is (well, we can ask, but we wouldn't likely get a clear answer!), its age needs to be estimated by other factors.  The variables in *dugong.csv* are length (in meters) and age (in years).  

Suppose we are interested in using the length of a dugong to predict its age.  We can fit a regression model for this!

Credit:  The *dugong.csv* file is from Data8 at UC-Berkeley.


### 1

- Read in the *dugong.csv* data set.  
-  Create a scatter plot with length on the x-axis and age on the y-axis; be sure to add descriptive axis labels (include units of measurement) and a title.  
-  Using `geom_smooth()`, add the least-squares line to your plot.
```{r}
dugong_orig=read_csv("dugong.csv")


ggplot(dugong_orig, aes(x = Length, y = Age)) +
  geom_point() +
  xlab("Length (Meters)") +
  ylab("Age (Years)") +
  ggtitle("Ages and Length of Dugongs") +
  geom_smooth(se = FALSE, method = "lm") +
  theme_bw()
```


### 2

- Using the dugong data, calculate the sample means, sample standard deviations, and correlation coefficient of the varaibles age and length.
- Using formulas from lecture, calculate the slope and intercept of the least squares regressions line to predict age with length.


```{r}
dugong_sum = dugong_orig %>% 
  summarize(across(everything(), list(mean = mean, sd = sd)),
            n = n(),
            r = cor(Length, Age)) %>% 
  relocate(n)

dugong_sum = dugong_sum %>% 
  mutate(slope_B1= r*(Age_sd/Length_sd)) %>% 
  mutate(intercept_B0= Age_mean - slope_B1*Length_mean)

dugong_sum %>% 
  print(widt=Inf)

```


### 3

- Use the dugong data and the functions `lm()` and `coef()` to calculate the slope and intercept of the least squares regression line of age against length (use length to predict age).
- How do the estimates using the two methods compare?


```{r}
dugong_lm = lm(Age ~ Length, data = dugong_orig)
cf = coef(dugong_lm)
cf
summary(dugong_lm)


```

> The estimates using either method obtain the same outputs for the intercept and mean.


### 4

- Add columns with the predicted values and residuals to the dugong data set. *(You can use* **modelr** *functions or just use `mutate()` and calculate these values directly.)*
- What are the mean and the standard deviation of the residuals?

```{r}
lm1 <- lm(Age~Length, data = dugong_orig)
summary(lm1)

dugong_orig <- dugong_orig%>%
  add_residuals(lm1) %>%
  add_predictions(lm1)
dugong_orig

dugong_mu_stddev = dugong_orig %>% 
  summarize(mean= mean(resid),
            std_dev= sd(resid))
dugong_mu_stddev
```


> the mean of the residuals= 1.552674e-14 and the standard deviation of the residuals= 4.393461	

### 5

- Plot the residuals versus length.
- Add to this plot a horizontal dashed red line with y intercept 0 and a smooth blue curve using `geom_smooth()` with no ribbon
- Add descriptive labels and a title.
- Comment on the appropriateness of a linear model to describe the relationship between length and age in dugongs.

```{r}
ggplot(dugong_orig, aes(x = Length, y = resid)) +
  geom_point() +
  xlab("Length (Meters)") +
  ylab("Residuals") +
  ggtitle("Residuals and Length of Dugongs")+
  geom_smooth(se=FALSE) +
  geom_hline(yintercept = 0, color = "red", linetype= "dashed")
```

>The two models for length and age in dugong is quite different, showing that a linear model is not an appropiate model for the relationship of the two variables because the residual are negative and positive and do not cluster together well.


### 6

- The simple linear regression model for $Y_i$ conditional on the values of $X_i = x_i$ is

$$
\E(Y_i \mid X_i = x_i) = \beta_0 + \beta_1 x_i + \varepsilon_i, \quad \text{for $i = 1, \ldots,n$}
$$

where $\varepsilon_i \sim \text{Normal}(0, \sigma)$
for some parameter $\sigma > 0$.

- The parameter $\sigma$ is the unknown population standard deviation of the typical distance between a point $Y_i$ and its true expected value.
- We can use the residuals, distances between the observed $y_i$ and the fitted regression line as an estimate of $\sigma$.
- However, the conventional estimate is **not** simply the standard deviation of the residuals, but is calculated by a very similar formula.

$$
\hat{\sigma} = \sqrt{ \frac{\sum_{i=1}^n (r_i - \bar{r})^2}{c} }
$$

where $r_i$ is the $i$th residual, $\bar{r}$ is the mean of the residuals (what is it equal to?), and $c$ is a number related to the sample size $n$ for you to determine.

- Use `lm()` to fit the regression line of age on length.
- Use  `summary()` on this fitted model object and read the results to find the numerical value of the estimate of $\sigma$, $\hat{\sigma}$.
  - Alternatively, there is a base R function named `sigma()` you can use to extract this value from a fitted `lm()` object.
  - Note, if you have a local variable named `sigma`, you would need to call the function with its prefix, `stats::sigma()`.
- Compare this value to the standard deviation of the residuals.
- By calculation or trial and error, what value of $c$ is needed in the equation above to replicate the value of $\hat{\sigma}$ for the regression model? Show your calculation to verify your response.

```{r}
summary(lm1)

```

$$
\hat{\sigma} = 4.48
$$
$$
{\sigma} = 4.39
$$
$$
c= 25
$$

> A value of 25 is required for c to replicate the regression model for sigma_hat.


### 7

- Read in the Boston marathon data from the file `boston-marathon-data.csv`.
- Create scatter plots of `Time` versus `Age` separately for each `Sex` by using a single call to `ggplot()` and using separate facets for each sex.
- Add a straight regression line to each plot and a smooth curve using `geom_smooth()` and no ribbon.
- Make two residual plots, one for each sex.
- Based on visual examination of these plots, is it reasonable to model `Time` versus `Age` with simple linear regression for each sex? Briefly explain.


```{r}
bos_mar=read_csv("boston-marathon-data.csv")

ggplot(bos_mar, aes(x = Age, y = Time)) +
  geom_point() +
  xlab("Age (Years)") +
  ylab("Time (Hours)") +
  ggtitle("Time and Age of Boston Marathon Finishers") +
  geom_smooth(se = FALSE, method = "lm") +
  geom_smooth(se=FALSE)+
  facet_wrap(~Sex)


lm2 <- lm(Time~Age, data = bos_mar)
bos_mar <- bos_mar%>%
  add_residuals(lm2)


ggplot(bos_mar, aes(x=Age, y=resid)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm")+
  xlab("Age (Years)") +
  ylab("Residuals") +
  ggtitle("Boston Marathon (Time vs Age) Residual Plot")+
  geom_hline(aes(yintercept=0), color="red", linetype = "dashed")+
  facet_wrap(~Sex)
```

>Yes, you can model Time vs. Age of males and females by a simple linear regression because the residuals are positive and there is evidence of clustering of points.


